---
title: "snRNA combined sampels"
author: "Elise RĂ¸ttereng"
date: "2026-02-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

Trying all samples combined!


# Set seed

```{r set seed}

set.seed(14)

```

set.seed do not work on UMAP


# Install Packages

```{r install packages}

.libPaths(c(
  "C:/Users/elise/AppData/Local/R/win-library/4.5",
  "C:/Rlibs",
  "C:/Program Files/R/R-4.5.1/library"))

suppressPackageStartupMessages(library(scDblFinder))

library(tidyverse)
library(Seurat)
library(DropletUtils)
library(patchwork)
library(presto)
library(readxl)
library(biomaRt)
library(purrr)
library(janitor)
library(rcartocolor)
library(Matrix)
library(pheatmap)
library(harmony)
library(speckle)

```


# Load the Data

Use ReadMtx(). Tells Seurat explicitly which files to use. 

```{r load data}

sample1_data <- ReadMtx(
  mtx = "raw1/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw1/features.tsv.gz",
  cells = "raw1/barcodes.tsv.gz",
  feature.column = 1) #chooses gene name as row names

sample2_data <- ReadMtx(
  mtx = "raw2/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw2/features.tsv.gz",
  cells = "raw2/barcodes.tsv.gz",
  feature.column = 1)

sample3_data <- ReadMtx(
  mtx = "raw3/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw3/features.tsv.gz",
  cells = "raw3/barcodes.tsv.gz",
  feature.column = 1)

sample5_data <- ReadMtx(
  mtx = "raw5/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw5/features.tsv.gz",
  cells = "raw5/barcodes.tsv.gz",
  feature.column = 1)

sample7_data <- ReadMtx(
  mtx = "raw7/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw7/features.tsv.gz",
  cells = "raw7/barcodes.tsv.gz",
  feature.column = 1)

sample9_data <- ReadMtx(
  mtx = "raw9/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw9/features.tsv.gz",
  cells = "raw9/barcodes.tsv.gz",
  feature.column = 1)

sample10_data <- ReadMtx(
  mtx = "raw10/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw10/features.tsv.gz",
  cells = "raw10/barcodes.tsv.gz",
  feature.column = 1)

sample11_data <- ReadMtx(
  mtx = "raw11/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw11/features.tsv.gz",
  cells = "raw11/barcodes.tsv.gz",
  feature.column = 1)

sample12_data <- ReadMtx(
  mtx = "raw12/UniqueAndMult-PropUnique.mtx.gz",
  features = "raw12/features.tsv.gz",
  cells = "raw12/barcodes.tsv.gz",
  feature.column = 1)

```

Special matrix where "0" are "." instead to make the object smaller.

Can run e.g. "data_set[1000:1020, 1000:1020]" to see


# Gather samples

```{r gather samples}

gc()

# Put samples into a list
samples <- list(
  sample1 = sample1_data,
  sample2 = sample2_data,
  sample3 = sample3_data,
  sample5 = sample5_data,
  sample7 = sample7_data,
  sample9 = sample9_data,
  sample10 = sample10_data,
  sample11 = sample11_data,
  sample12 = sample12_data)

```


# Sum of all counts for every gene

Kinda remove the cells, so that I can compare with the bulk RNA

```{r sum counts}

# Make function
sum_per_gene <- function(mat) {
  rowSums(mat)
}

# Sum counts for every gene
sum_list <- lapply(samples, sum_per_gene)

# Make into a matrix
sum_matrix <- do.call(cbind, sum_list)

```


# Save and Download

Save as R object

```{r save}

save(sum_matrix, file = "all_couts_genes_object.RData")

```

Downlaod

```{r load}

load(file = "all_couts_genes_object.RData")

```

Clean up

```{r remove}

rm(sum_per_gene, sum_list, sum_matrix)

```


# Remove empty droplets and low counts

This needs to be done before we can remove doublets

Get bar code information.

Make a function that gets the information from all samples

```{r information funk}

# Function that gets the knee and inflection for counts and features
barcode_funk <- function(x){
  
  # Counts
  br_counts <- barcodeRanks(x)
  
  # Features
  
  br_features <- barcodeRanks(x > 0)
  
  # Knee and inflection
  
  list(
    counts = list(
      br = br_counts,
      knee = metadata(br_counts)$knee,
      inflection = metadata(br_counts)$inflection),
    features = list(
      br = br_features,
      knee = metadata(br_features)$knee,
      inflection = metadata(br_features)$inflection))
}

```

Run the function

```{r run funk}

gc()

barcode_info <- lapply(samples, barcode_funk)

```

Make a function that plot the barcodes and their cutoff values

```{r plot function}

plot_barcode <- function(br_obj, ylab, main_title) {
  
  # Get knee and inflection
  
  knee <- metadata(br_obj)$knee
  inflection <- metadata(br_obj)$inflection
  
  # Convert to data frame and sort
  
  df <- as.data.frame(br_obj)
  df <- df[order(df$rank), ]
  
  # Remove empty barcodes
  
  df <- df[df$total > 0, ]
  
  # Plot
  
  plot(df$rank,
       df$total,
       log = "xy",
       xlab = "Cell rank",
       ylab = ylab,
       pch = 16,
       col = "#1D6996",
       main = main_title)
  
  if (!is.null(inflection))
    abline(h = inflection, col = "grey", lty = 2)
  
  if (!is.null(knee))
    abline(h = knee, col = "black", lty = 2)
  
  recordPlot()
}
```

Plot

```{r plot}

gc()

# Sample 1
barcode1_features_plot <- plot_barcode(
  br_obj = barcode_info$sample1$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 1 - Features")

# Sample 2
barcode2_features_plot <- plot_barcode(
  br_obj = barcode_info$sample2$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 2 - Features")

# Sample 3
barcode3_features_plot <- plot_barcode(
  br_obj = barcode_info$sample3$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 3 - Features")

# Sample 5
barcode5_features_plot <- plot_barcode(
  br_obj = barcode_info$sample5$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 5 - Features")

# Sample 7
barcode7_features_plot <- plot_barcode(
  br_obj = barcode_info$sample7$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 7 - Features")

# Sample 9
barcode9_features_plot <- plot_barcode(
  br_obj = barcode_info$sample9$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 9 - Features")

# Sample 10
barcode10_features_plot <- plot_barcode(
  br_obj = barcode_info$sample10$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 10 - Features")

# Sample 11
barcode11_features_plot <- plot_barcode(
  br_obj = barcode_info$sample11$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 11 - Features")

# Sample 12
barcode12_features_plot <- plot_barcode(
  br_obj = barcode_info$sample12$features$br,
  ylab = "Features",
  main_title = "Barcode Rank 12 - Features")

```

Decide which barcodes to keep

Keeping the inflection values on the features

```{r keep}

gc()

# Keeping barcodes

# Sample 1
br1_features <- barcode_info$sample1$features$br
inflection1 <- barcode_info$sample1$features$inflection

keep_barcodes1 <- colnames(sample1_data)[br1_features$total >= inflection1]

# Sample 2
br2_features <- barcode_info$sample2$features$br
inflection2 <- barcode_info$sample2$features$inflection

keep_barcodes2 <- colnames(sample2_data)[br2_features$total >= inflection2]

# Sample 3
br3_features <- barcode_info$sample3$features$br
inflection3 <- barcode_info$sample3$features$inflection

keep_barcodes3 <- colnames(sample3_data)[br3_features$total >= inflection3]

# Sample 5
br5_features <- barcode_info$sample5$features$br
inflection5 <- barcode_info$sample5$features$inflection

keep_barcodes5 <- colnames(sample5_data)[br5_features$total >= inflection5]

# Sample 7
br7_features <- barcode_info$sample7$features$br
inflection7 <- barcode_info$sample7$features$inflection

keep_barcodes7 <- colnames(sample7_data)[br7_features$total >= inflection7]

# Sample 9
br9_features <- barcode_info$sample9$features$br
inflection9 <- barcode_info$sample9$features$inflection

keep_barcodes9 <- colnames(sample9_data)[br9_features$total >= inflection9]

# Sample 10
br10_features <- barcode_info$sample10$features$br
inflection10 <- barcode_info$sample10$features$inflection

keep_barcodes10 <- colnames(sample10_data)[br10_features$total >= inflection10]

# Sample 11
br11_features <- barcode_info$sample11$features$br
inflection11 <- barcode_info$sample11$features$inflection

keep_barcodes11 <- colnames(sample11_data)[br11_features$total >= inflection11]

# Sample 12
br12_features <- barcode_info$sample12$features$br
inflection12 <- barcode_info$sample12$features$inflection

keep_barcodes12 <- colnames(sample12_data)[br12_features$total >= inflection12]

```

Remove objects that are no longer useful

```{r clean up}

rm(barcode_funk, plot_barcode, inflection1, inflection2, inflection3, inflection5, inflection7, inflection9, inflection10, inflection11, inflection12, br1_features, br2_features, br3_features, br5_features, br7_features, br9_features, br10_features, br11_features, br12_features, samples)

```


# Make Seurat Object

After having removed the empty droplets, do not used the min.cells or the min.features, because they have already been filtered out.

```{r seurat object}

gc()

seu1 <- CreateSeuratObject(
  counts = sample1_data[, keep_barcodes1],
  project = "sample1")

seu2 <- CreateSeuratObject(
  counts = sample2_data[, keep_barcodes2],
  project = "sample2")

seu3 <- CreateSeuratObject(
  counts = sample3_data[, keep_barcodes3],
  project = "sample3")

seu5 <- CreateSeuratObject(
  counts = sample5_data[, keep_barcodes5],
  project = "sample5")

seu7 <- CreateSeuratObject(
  counts = sample7_data[, keep_barcodes7],
  project = "sample7")

seu9 <- CreateSeuratObject(
  counts = sample9_data[, keep_barcodes9],
  project = "sample9")

seu10 <- CreateSeuratObject(
  counts = sample10_data[, keep_barcodes10],
  project = "sample10")

seu11 <- CreateSeuratObject(
  counts = sample11_data[, keep_barcodes11],
  project = "sample11")

seu12 <- CreateSeuratObject(
  counts = sample12_data[, keep_barcodes12],
  project = "sample12")

```

Remove unnecessary objects 

```{r clean up}

rm(sample1_data, sample2_data, sample3_data, sample5_data, sample7_data, sample9_data, sample10_data, sample11_data, sample12_data, keep_barcodes1, keep_barcodes2, keep_barcodes3, keep_barcodes5, keep_barcodes7, keep_barcodes9, keep_barcodes10, keep_barcodes11, keep_barcodes12)

```

 
# Removing doublets

Prepare the inputs: a SingleCellExperiment object

```{r make SCE object}

gc()

sce1 <- as.SingleCellExperiment(seu1)

sce2 <- as.SingleCellExperiment(seu2)

sce3 <- as.SingleCellExperiment(seu3)

sce5 <- as.SingleCellExperiment(seu5)

sce7 <- as.SingleCellExperiment(seu7)

sce9 <- as.SingleCellExperiment(seu9)

sce10 <- as.SingleCellExperiment(seu10)

sce11 <- as.SingleCellExperiment(seu11)

sce12 <- as.SingleCellExperiment(seu12)

```

Run scDoubletFinder

```{r run dobletFinder}

gc()

sce1 <- scDblFinder(sce1) #Can add expected number of doublets

sce2 <- scDblFinder(sce2) 

sce3 <- scDblFinder(sce3) 

sce5 <- scDblFinder(sce5)

sce7 <- scDblFinder(sce7)

sce9 <- scDblFinder(sce9)

sce10 <- scDblFinder(sce10)

sce11 <- scDblFinder(sce11)

sce12 <- scDblFinder(sce12)

```

Add the labels back to the Seurat object

Need to add the cell_IDs as row names

Extract the metadata

```{r extract metadata}

gc()

meta_scdblfinder1 <- sce1@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder2 <- sce2@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder3 <- sce3@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder5 <- sce5@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder7 <- sce7@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder9 <- sce9@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder10 <- sce10@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder11 <- sce11@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

meta_scdblfinder12 <- sce12@colData@listData %>% as.data.frame() %>% dplyr::select(starts_with('scDblFinder'))

```

Change row names to cell-IDs

```{r change rownames}

rownames(meta_scdblfinder1) <- sce1@colData@rownames

rownames(meta_scdblfinder2) <- sce2@colData@rownames

rownames(meta_scdblfinder3) <- sce3@colData@rownames

rownames(meta_scdblfinder5) <- sce5@colData@rownames

rownames(meta_scdblfinder7) <- sce7@colData@rownames

rownames(meta_scdblfinder9) <- sce9@colData@rownames

rownames(meta_scdblfinder10) <- sce10@colData@rownames

rownames(meta_scdblfinder11) <- sce11@colData@rownames

rownames(meta_scdblfinder12) <- sce12@colData@rownames

```

Add metadata to the seurat object

```{r meta seurat}

gc()

seu1 <- AddMetaData(object = seu1, metadata = meta_scdblfinder1 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score')) 

seu2 <- AddMetaData(object = seu2, metadata = meta_scdblfinder2 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score')) 

seu3 <- AddMetaData(object = seu3, metadata = meta_scdblfinder3 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score')) 

seu5 <- AddMetaData(object = seu5, metadata = meta_scdblfinder5 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score'))

seu7 <- AddMetaData(object = seu7, metadata = meta_scdblfinder7 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score'))

seu9 <- AddMetaData(object = seu9, metadata = meta_scdblfinder9 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score'))

seu10 <- AddMetaData(object = seu10, metadata = meta_scdblfinder10 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score'))

seu11 <- AddMetaData(object = seu11, metadata = meta_scdblfinder11 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score'))

seu12 <- AddMetaData(object = seu12, metadata = meta_scdblfinder12 %>% dplyr::select('scDblFinder.class', 'scDblFinder.score'))

```

Remove objects

```{r clean}

# Remove temporary variables to free up memory 
rm(meta_scdblfinder1, meta_scdblfinder2, meta_scdblfinder3, meta_scdblfinder5, meta_scdblfinder7, meta_scdblfinder9, meta_scdblfinder10, meta_scdblfinder11, meta_scdblfinder12, sce1, sce2, sce3, sce5, sce7, sce9, sce10, sce11, sce12)

```

Remove the doublets

```{r remove}

gc()

seu1 <- subset(seu1, subset = scDblFinder.class == 'singlet') 

seu2 <- subset(seu2, subset = scDblFinder.class == 'singlet') 

seu3 <- subset(seu3, subset = scDblFinder.class == 'singlet') 

seu5 <- subset(seu5, subset = scDblFinder.class == 'singlet') 

seu7 <- subset(seu7, subset = scDblFinder.class == 'singlet') 

seu9 <- subset(seu9, subset = scDblFinder.class == 'singlet') 

seu10 <- subset(seu10, subset = scDblFinder.class == 'singlet') 

seu11 <- subset(seu11, subset = scDblFinder.class == 'singlet') 

seu12 <- subset(seu12, subset = scDblFinder.class == 'singlet') 

```


# Identify and removing Mitochondria

Identify

```{r meta data}

gc()

seu1[["percent.mt"]] <- PercentageFeatureSet(seu1, pattern = "^MT-")

seu2[["percent.mt"]] <- PercentageFeatureSet(seu2, pattern = "^MT-")

seu3[["percent.mt"]] <- PercentageFeatureSet(seu3, pattern = "^MT-")

seu5[["percent.mt"]] <- PercentageFeatureSet(seu5, pattern = "^MT-")

seu7[["percent.mt"]] <- PercentageFeatureSet(seu7, pattern = "^MT-")

seu9[["percent.mt"]] <- PercentageFeatureSet(seu9, pattern = "^MT-")

seu10[["percent.mt"]] <- PercentageFeatureSet(seu10, pattern = "^MT-")

seu11[["percent.mt"]] <- PercentageFeatureSet(seu11, pattern = "^MT-")

seu12[["percent.mt"]] <- PercentageFeatureSet(seu12, pattern = "^MT-")

```

Removing 

```{r removing}

gc()

seu1 <- subset(seu1, subset = percent.mt < 1)

seu2 <- subset(seu2, subset = percent.mt < 1)

seu3 <- subset(seu3, subset = percent.mt < 1)

seu5 <- subset(seu5, subset = percent.mt < 1)

seu7 <- subset(seu7, subset = percent.mt < 1)

seu9 <- subset(seu9, subset = percent.mt < 1)

seu10 <- subset(seu10, subset = percent.mt < 1)

seu11 <- subset(seu11, subset = percent.mt < 1)

seu12 <- subset(seu12, subset = percent.mt < 1)

```


# Merging the samples

```{r merge}

seu_merged <- merge(seu1, y = c(seu2, seu3, seu5, seu7, seu9, seu10, seu11, seu12), 
                    add.cell.ids = c("sample1", "sample2", "sample3", "sample5", "sample7", "sample9", "sample10", "sample11", "sample12"),
                    project = "sampleIntegrated")

```

Amount of cells per sample

```{r look}

table(seu_merged$orig.ident)

seu_merged$orig.ident <- factor(
  seu_merged$orig.ident,
  levels = c("sample1", "sample2", "sample3", "sample5",
             "sample7", "sample9", "sample10",
             "sample11", "sample12"))

cellnumber_df <- as.data.frame(table(seu_merged$orig.ident))
colnames(cellnumber_df) <- c("Sample", "Cells")

ggplot(cellnumber_df, aes(x = Sample, y = Cells, fill = Sample)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Sample", y = "Number of cells") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none")

```

Remove objects

```{r clean up}
rm(seu1, seu2, seu3, seu5, seu7, seu9, seu10, seu11, seu12)
```


# Save and Downlaod 

```{r save}

save(seu_merged, file = "merged_object.RData")

```

```{r Load}

load(file = "merged_object.RData")

```



# Normalizing the Data

```{r normalizing data}

seu_merged <- NormalizeData(seu_merged, normalization.method = "LogNormalize", scale.factor = 10000)

```


# Identification of Highly Variable Features (feature selection)

```{r variable features}

seu_merged <- FindVariableFeatures(seu_merged, selection.method = "vst", nfeatures = 2000)

```


# Scaling

Linear transformation - makes mean expression 0, scale expression so variance across cells is 1 (highly expressed genes do not dominate), and more. Remove unwanted source of variation (e.g technical variation, variation because of cell cycle)

```{r scaling}

gc()

seu_merged <- ScaleData(seu_merged)

```


# Linear Dimension Reduction

Preform PCA. 

```{r PCA}

gc()

seu_merged <- RunPCA(seu_merged)

```

Make elbow plot

```{r elbow plot}

gc()

elbow_plot <- ElbowPlot(seu_merged)

elbow_plot

```


# Cluster the Cells

Uses K-nearest neighbor using euclidean distance in the PCA space, and Jaccard similarity.

```{r cluster cells}

# Find Neighbors
seu_merged <- FindNeighbors(seu_merged, dims = 1:12, reduction = "pca")

# Find Clusters
seu_merged <- FindClusters(seu_merged, resolution = 0.4, cluster.name = "mergedClusters")

head(Idents(seu_merged), 5)

```


# Non-Linear Dimention Reductrion (UMAP)

```{r UMAP}

gc()

seu_merged <- RunUMAP(seu_merged, dims = 1:12, reduction = "pca", reduction.name = "mergedUMAP")

```


# Save and laod

```{r save}

save(seu_merged, file = "merged_UMAP_object.RData")

```

```{r Load}

load(file = "merged_UMAP_object.RData")

```

# Plot

Plot

```{r plot UMAP}

seu_merged$orig.ident <- factor(
  seu_merged$orig.ident,
  levels = c("sample1", "sample2", "sample3", "sample5",
             "sample7", "sample9", "sample10",
             "sample11", "sample12"))

merged_umap_plot <- DimPlot(seu_merged, reduction = "mergedUMAP", group.by = "orig.ident") + 
  NoAxes() +
  theme_void()
  
merged_umap_plot

```


# Sample Integration

```{r integration}

seu_integrated <- IntegrateLayers(object = seu_merged, method = HarmonyIntegration, orig.reduction = "pca", new.reduction = "harmony", verbose = T)

```


# Save and laod

```{r save}

save(seu_integrated, file = "integrated_object.RData")

```

```{r Load}

load(file = "integrated_object.RData")

```


# Find clusters

Find neighbors 

```{r neighbors}

gc()

seu_integrated <- FindNeighbors(seu_integrated, reduction = "harmony", dims = 1:12)

```

Cluster

```{r clusters}

seu_integrated <- FindClusters(seu_integrated, resolution = 0.4, cluster.name = "integratedClusters")

```


# Run UMAP

```{r UMAP}

gc()

seu_integrated <- RunUMAP(seu_integrated, reduction = "harmony", dims = 1:12, reduction.name = "integratedUMAP")

```


# Save and laod

```{r save}

save(seu_integrated, file = "integrated_UMAP_object.RData")

```

Load

```{r Load}

load(file = "integrated_UMAP_object.RData")

```

# Plot

```{r plot}

integrated_umap_plot <- DimPlot(seu_integrated, reduction = "integratedUMAP", group.by = "orig.ident") + 
  NoAxes() +
  theme_void()

integrated_umap_plot

```

Plot UMAP with cluster number

```{r plot cluster}

integrated_cluster_umap_plot <- DimPlot(seu_integrated, reduction = "integratedUMAP", label = T) + 
  NoAxes() +
  theme_void()

integrated_cluster_umap_plot

```

Expression

```{r expresson}

FeaturePlot(seu_integrated, features = "nFeature_RNA", reduction = "integratedUMAP", min.cutoff = "q05", max.cutoff = "q95") + 
  NoAxes() +
  theme_void() 

```

Doublet score

```{r doublet score }

FeaturePlot(seu_integrated, features = 'scDblFinder.score')

```

Look at where there is mitochondria gene expression, it is not suppose to be here. Look where they have maybe attached themselves to nuclei. Mess with things. 

```{r mitochondria}

FeaturePlot(seu_integrated, features = "MT-CYTB")

```

Better than FeaturePlot function

```{r function}

# Color pallet
sccol <- (colorRampPalette(c(carto_pal(7, "Sunset"))))(50) #gene expression color scale


# FeaturePlot-like for a Seurat v5 object:

# scplot5("nCount_RNA", seuratobject, sccol)

# scplot5(expressionvalues, seuratobject, sccol, extradata = T)

# Always plots the most interesting cells (with high expression) over less interesting cells :)



#diverging = F makes color gradient, pch = points, cex = point size, add = F makes a new plot, extradata for when the genes are not gene names

scplot5 <- function(gene, dataset, xcol, diverging = FALSE, pch = 16,
                    cex = 1, add = FALSE, extradata = FALSE,
                    log = FALSE, main = TRUE) {
  
  maxscale <- function(x) x / max(abs(x), na.rm = TRUE)
  
  # If more than one gene
  
  if (length(gene) > 1 && extradata == FALSE) {

    oldpar <- par(no.readonly = TRUE)
    on.exit(par(oldpar))

    n <- length(gene)
    rows <- floor(sqrt(n))
    cols <- ceiling(n / rows)
    par(mfrow = c(rows, cols), mar = c(1, 1, 3, 1))

    for (g in gene) {
      scplot5(
        gene = g,
        dataset = dataset,
        xcol = xcol,
        diverging = diverging,
        pch = pch,
        cex = cex,
        add = FALSE,
        extradata = FALSE,
        log = log,
        main = TRUE
      )
    }
    return(invisible(NULL))
  }

  # Handling one gene at the time
  
  if (extradata == FALSE) {
    x <- as.numeric(FetchData(dataset, gene)[,1])} 
  else {
    x <- gene
    gene <- deparse(substitute(gene))}

  
  if (log == TRUE) {
    x <- log1p(x)}
  
  y <- order(abs(x)) #strongest expression last

  
  if (diverging == TRUE) {
    x_scaled <- maxscale(x)
    ycol <- xcol[
      ceiling(0.00001 + (x_scaled[y] + 1) *(length(xcol) - 1) * 0.5)]} 
  else {
    ycol <- xcol[1 + ceiling((length(xcol) - 1) *((x[y] - min(x))/(max(x) - min(x))))]}
  
  # Get coordinates
  umap1 <- dataset@reductions$integratedUMAP@cell.embeddings[,1]
  umap2 <- dataset@reductions$integratedUMAP@cell.embeddings[,2]
  
  # Plot
  plot(
    umap1[y], umap2[y],
    col = ycol, pch = pch, cex = cex,
    main = gene,
    xlab = "", ylab = "",
    xaxt = "n", yaxt = "n",
    cex.main = 1.6
  )
}

```

Hub-locus genes

CTSZ

```{r plot}

scplot5("ENSSSAG00000005903", seu_integrated, sccol)

```

MPPE1

```{r plot}

scplot5("ENSSSAG00000052122", seu_integrated, sccol)

```

GNAI1

```{r plot}

scplot5("ENSSSAG00000099904", seu_integrated, sccol)

```

pak1Ip1

```{r plot}

scplot5("ENSSSAG00000004575", seu_integrated, sccol)

```

endoU

```{r plot}

scplot5("ENSSSAG00000004590", seu_integrated, sccol)

```

FAM111A

```{r plot}

scplot5("ENSSSAG00000004608", seu_integrated, sccol)

```

ZFYVE9

```{r plot}

scplot5("ENSSSAG00000007619", seu_integrated, sccol)

```


# Join Layers

```{r layers}

gc()

seu_integrated <- JoinLayers(seu_integrated)

```


# Save and load

```{r save}

save(seu_integrated, file = "join_layers_object.RData")

```

```{r Load}

load(file = "join_layers_object.RData")

```


# Cluster annotation

## Finding Differntially Expressed Features (cluster biomarkers)

Markers for every cluster compared to the remaining cells

Report only positive ones.

```{r markers}

gc()

seu_markers <- FindAllMarkers(seu_integrated, only.pos = TRUE) %>% 
  dplyr::filter(avg_log2FC > 1)

seu_cluster_markers <- with(seu_markers, lapply(split(gene, cluster), unique))

```

Remove objects

```{r clean }

rm(seu_markers)

```


# Save and load markers

```{r save markers}

save(seu_cluster_markers, file = "cluseter_markers_object.RData")

```

```{r Load markers}

load(file = "cluseter_markers_object.RData")

```

## Known Cell Type Marker Genes

Download genes

```{r donwlaod Ensembl IDs}

mart <- useEnsembl(
  biomart = "ensembl",
  dataset = "ssalar_gene_ensembl")

```

Make a function to extract the markers from the excel file

```{r Make function}

process_excel_with_biomaRt <- function(file, mart) {

  # Find all sheets 
  sheets <- excel_sheets(file)

  # Read all sheets and put them in one table
  all_markers <- map_dfr(sheets, function(sheet_name) { #make a loop
    df <- read_excel(file, sheet = sheet_name) %>%  #read sheet
      clean_names()  #clean up name

    if (!"gene_id" %in% names(df)) { #skip sheets without gene_id ergo GO sheets
      return(NULL)
    }

    df %>%
      mutate(original_order = row_number()) %>% 
      separate(gene_id, into = c("Gene", "NCBI_ID"), sep = ":") %>% 
      mutate(
        NCBI_ID = as.character(NCBI_ID), 
        sheet = sheet_name) %>% 
      dplyr::select(sheet, NCBI_ID, original_order)
    })
  
  # Remove duplicates and NA
  all_markers <- filter(all_markers, !is.na(NCBI_ID))
  
  unique_ids <- unique(all_markers$NCBI_ID)

  # BiomaRt conversion
  conversion <- getBM( 
    attributes = c("entrezgene_id", "ensembl_gene_id"),
    filters = "entrezgene_id",
    values = unique(all_markers$NCBI_ID),
    mart = mart)
  
  conversion$entrezgene_id <- as.character(conversion$entrezgene_id)

  # Put the ensembl IDs into the table 
  merged <- merge(
    all_markers,
    conversion,
    by.x = "NCBI_ID",
    by.y = "entrezgene_id",
    all.x = T)
  
  merged <- merged[!is.na(merged$ensembl_gene_id), ]
  
  merged <- merged[order(merged$sheet, merged$original_order), ]
  
  marker_list <- lapply(split(
    merged$ensembl_gene_id,
    merged$sheet),
    unique)
  
  return(marker_list)
}

```

Run Function

```{r run function}

marker_list <- process_excel_with_biomaRt("Markergenes.XLSX", mart)

rm(mart)
  
```


# Save and Load

save

```{r save}

save(marker_list, file = "known_markers.Rdata")

```

load

```{r load}

load(file = "known_markers.Rdata")

```


## Find overlap

Make a matrix that counts the overlap for every cluster and cell type

```{r overlap}

overlap_clust_known <- sapply(marker_list, function(celltypes_genes) {
  sapply(seu_cluster_markers, function(cluster_genes) {
    length(intersect(cluster_genes, celltypes_genes))
    })
  })

overlap_clust_known

```

Heatmap of Overlap

```{r heatmap}

pheatmap(
  overlap_clust_known,
  color = colorRampPalette(c("white", "blue")) (50), #50 different colors
  cluster_rows = T, #order based on overlap
  cluster_cols = T,
  treeheight_row = 0, #remove lines
  treeheight_col = 0,
  display_numbers = F) #do not have numbers
```


## Percent overlap

Calculate by dividing on the number of marker genes found for every cluster

```{r % devided on cluster genes}

overlap_fraction_cluster <- sapply( #loop every cell type
  marker_list, 
  function(celltype_genes) { #one list of genes for every cell type - column 
    sapply(seu_cluster_markers, #loop every cluster
           function(cluster_genes) { #one list of  genes for every cluster - row
             length(intersect(cluster_genes, celltype_genes))/length(cluster_genes)
             }) #find and count all genes that are in both lists, divide with number of marker genes in the cluster
    })

overlap_fraction_cluster

```

Plot heatmap

```{r heatmap}

pheatmap(
  overlap_fraction_cluster,
  color = colorRampPalette(c("white", "red")) (50),
  cluster_rows = T, 
  cluster_cols = T,
  treeheight_row = 0,
  treeheight_col = 0,
  display_numbers = F)

```

Calculate by dividing on the number of marker genes found for every cluster

```{r % divided by marker genes}

overlap_fraction_celltype_cluster <- sapply( #loop every cell type
  marker_list, 
  function(celltype_genes) { #one list of genes for every cell type - column
    sapply(seu_cluster_markers, #loop every cluster
           function(cluster_genes) { #one list of  genes for every cluster - row
             length(intersect(cluster_genes, celltype_genes))/length(celltype_genes)
             }) #find and count all genes that are in both lists, divide with number of known marker genes
    })

overlap_fraction_celltype_cluster

```

Plot heatmap

```{r heatmap}

pheatmap(
  overlap_fraction_celltype_cluster,
  color = colorRampPalette(c("white", "purple")) (50),
  cluster_rows = T, 
  cluster_cols = T,
  treeheight_row = 0,
  treeheight_col = 0,
  display_numbers = F)

```

## Annotation

```{r annotation}

# Apply is used for matrix, 1 means work row for row, every row is one cluster.

cluster_annotation <- apply(overlap_clust_known, 1, function(x) {
  if (max(x) < 0.05) { #if the highest % overlap is 0.05, call it uncertain 
    "Uncertain"
  } else { #else call it by the name of the cell type with the highest overlap
    names(which.max(x))
  }})

cluster_annotation

# Add to seu_integrated

new_labels <- cluster_annotation[as.character(seu_integrated$integratedClusters)]
names(new_labels) <- colnames(seu_integrated)

seu_integrated$celltype_annotation <- new_labels

```


# Save and loade

The annotated Seurat object

save

```{r save}

save(seu_integrated, file = "annotated_seurat.Rdata")

```

load

```{r load}

load(file = "annotated_seurat.Rdata")

```


# Plot UMAP annotated

```{r plot}

umap_anot_plot <- DimPlot(seu_integrated, reduction = "integratedUMAP", group.by = "celltype_annotation", label = TRUE, repel = TRUE)+ 
  NoAxes() +
  theme_void()

umap_anot_plot

```

Put the genotype in to the metadata

```{r genotype to meta}

cc_samples <- c("sample1", "sample2", "sample3", "sample5", "sample7", "sample10", "sample13", "sample14")

tt_samples <- c("sample9", "sample11", "sample12", "sample15", "sample16")


seu_integrated$genotype <- ifelse(
  seu_integrated$orig.ident %in% cc_samples, "CC",
  ifelse(seu_integrated$orig.ident %in% tt_samples, "TT", NA)) %>% 
  factor(levels = c("CC", "TT"))

# Set order

seu_integrated$orig.ident <- factor(
  seu_integrated$orig.ident,
  levels = c(cc_samples, tt_samples))

```


# Cell type composition

Make one CC plot and one TT plot

```{r genotype umap }

genotype_umap_plot <- DimPlot(seu_integrated, reduction = "integratedUMAP", group.by = "celltype_annotation", split.by = "genotype", label = T) + 
  NoAxes() +
  theme_void()

genotype_umap_plot 

```

Make split on samples

```{r all samples one plot}

all_samples_umap_plot <- DimPlot(seu_integrated, reduction = "integratedUMAP", group.by = "celltype_annotation", split.by = "orig.ident", ncol = 4) + 
  NoAxes() +
  theme_void()

all_samples_umap_plot

```

Plot all the samples in individual plots

```{r all samples seperate}

samples <- unique(seu_integrated$orig.ident)

xlims <- range(Embeddings(seu_integrated, "integratedUMAP")[,1])
ylims <- range(Embeddings(seu_integrated, "integratedUMAP")[,2])

for(s in samples) {
  p <- DimPlot(
    seu_integrated,
    reduction = "integratedUMAP", 
    group.by = "celltype_annotation", 
    cells = WhichCells(seu_integrated, expression = orig.ident == s)) + 
    ggtitle(s) + 
    xlim(xlims) + 
    ylim(ylims)
  
  print(p)

}


```

Plot one sample

```{r one sample at the time}

DimPlot(
  seu_integrated,
  reduction = "integratedUMAP",
  group.by = "celltype_annotation",
  cells = WhichCells(seu_integrated, expression = orig.ident == "sample9"))+
  xlim(xlims) + 
  ylim(ylims)

```


## Amount of cells per celltypes for each sample

```{r cells per celltype}

cells_in_celltype <- table(seu_integrated$celltype_annotation, seu_integrated$orig.ident)

cells_in_celltype

```


## Are the cell composition different based on genotype


### Propeller

Found this R function, used specifically to find difference in cell type composition in scRNA-seq data. 

Run propeller

```{r propeller}

propeller_results <- propeller(clusters = seu_integrated$celltype_annotation,
                          sample = seu_integrated$orig.ident,
                          group = seu_integrated$genotype)

propeller_results

```

No FDR under 0.05 - no difference based on genotype. 

```{r propeller_results}

write.csv(propeller_results, "propeller_results")

```


# Proportions

Calculate the proportions of every cell type

Plot one celltype and see the difference between genotypes

```{r plot}

cell_proportion <- seu_integrated@meta.data %>%
  count(orig.ident, genotype, celltype_annotation) %>%
  group_by(orig.ident) %>%
  mutate(Proportions = n/sum(n)) %>%
  mutate(Percent = n / sum(n) * 100) %>%
  ungroup() %>% 
  mutate(orig.ident = recode(orig.ident, 
                             "sample1"  = "1CC",
                             "sample2"  = "2CC",
                             "sample3"  = "3CC",
                             "sample5"  = "5CC",
                             "sample7"  = "7CC",
                             "sample10" = "10CC",
                             "sample9"  = "9TT",
                             "sample11" = "11TT",
                             "sample12" = "12TT")) %>% 
  mutate(orig.ident = factor(orig.ident, levels = c("1CC","2CC","3CC","5CC","7CC", "10CC", "9TT","11TT","12TT")))

cell_proportion %>%
  group_by(orig.ident) %>%
  summarise(total = sum(Proportions))

```

```{r save}

write.csv(cell_proportion, "cell_proportion")

```

Plot

```{r plot}

set.seed(14)
  
ggplot(subset(cell_proportion, celltype_annotation == "Pillar Cells"),
       aes(x = genotype, y = Proportions, colour = genotype)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("Pillar Cells")

```

Cell type proportions

```{r proportions plot}

ggplot(cell_proportion, aes(x = celltype_annotation, y = Proportions, color = celltype_annotation, shape = genotype)) +
  geom_point(size = 2.5, stroke = 1.5, position = position_dodge(width = 0.75)) +
  scale_shape_manual(values = c(20, 22)) +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1)) +
  labs(x = "Cell type", color = "Cell type", shape = "Genotype")

```

Scaled cell type proportions

```{r scaled plot}

cell_proportion_scaled <- cell_proportion %>%
  group_by(celltype_annotation) %>%
  mutate(Proportion_scaled =
           (Proportions - min(Proportions))/
           (max(Proportions) - min(Proportions)))

ggplot(cell_proportion_scaled, aes(x = celltype_annotation, y = Proportion_scaled, color = celltype_annotation, shape = genotype)) +
  geom_point(size = 2.5, stroke = 1.5, position = position_dodge(width = 0.75)) +
  scale_shape_manual(values = c(15, 22)) +
  theme_minimal() +
  theme(axis.text = element_text(angle = 45, hjust = 1)) +
  labs(x = "Cell type", color = "Cell type")


```

Proportions all samples

```{r prop plot}

ggplot(cell_proportion, aes(x = orig.ident, y = Proportions, fill = celltype_annotation)) +
  geom_bar(stat = "identity") +
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  labs(x = "Sample")


```

Heat map

First, make into a matrix

```{r martrix}

cell_prop_matrix <- cell_proportion %>% 
  dplyr::filter(!is.na(celltype_annotation)) %>% 
  dplyr::select(orig.ident, celltype_annotation, Percent) %>% 
  pivot_wider(names_from = orig.ident,
              values_from = Percent,
              values_fill = 0) %>% 
  column_to_rownames("celltype_annotation") %>% 
  as.matrix()

```

save

```{r save}

write.csv(cell_prop_matrix, "cell_prop_matrix")

```

Heatmap

```{r heatmap}

pheatmap(cell_prop_matrix,
         color = colorRampPalette(c("white", "violetred"))(100),
         cluster_rows = T,
         cluster_cols = F,
         treeheight_row = 0,
         treeheight_col = 0,
         border_color = "grey60",
         fontsize_row = 10,
         fontsize_col = 10,
         display_numbers = T)

```


